Bayesian estimation of chronic disease epidemiology from incomplete data

### 2021-10-07 metahit/paper_analyses.Rmd
Updated analysis scripts

foo <- function(x){
  cat(deparse(substitute(x)))
}

ht <- 3
pdf("../../write/mort_data.pdf", height=ht)
gbdplot$areagender <- interaction(gbdplot$area, gbdplot$gender)
ggplot(gbdplot, aes(x=age, y=mort_prob, col=area, lty=gender, group=areagender)) + 
  facet_wrap(~disease, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual mortality probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

@@

## Convergence checks.   All OK with 5000 samples and the occasional model simplification
resnc <- resnh %>% filter(Rhat > 1.05)
table(resnc$disease, resnc$area, resnc$gender)

## Compare modal approx with posterior 
## Page of plots for a single area and gender.  Disease variation more important than area/gender variation  
## stmc, utrc use national point estimates 
## What are the current settings used for these?

resall %>% 
  filter(var=="cf", disease=="Dementia", gender=="Female", 
         area=="Bristol", between(age, 70, 90), model=="Independent areas")

resoptunc <- readRDS("resoptunc_nonhier.rds") %>% 
  filter(!disease %in% c("Stomach cancer","Uterine cancer"))
resoptnat <- readRDS("resopt_nat.rds")
resoptunccf <- resoptunc %>%
  full_join(resoptnat) %>% 
  filter(var=="cf") %>%
  mutate(p50 = 1 - exp(-`par`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  filter(area=="Bristol" | area=="England", gender=="Female") %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten),
           disease = factor(disease, levels = disease_order)) %>%
  filter(age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(Algorithm = "Mode approximation")

@@

resp <- resall %>% 
  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
  dplyr::filter(area=="Bristol" | (area=="England" & model=="National"), gender=="Female") %>%
  mutate(Algorithm="MCMC") %>%
  full_join(resoptunccf)
resp %>%
  filter(age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(par = 1 - exp(-`50%`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  ggplot(aes(x=age, y=par, col=algorithm, lty=algorithm)) + 
  geom_errorbar(aes(ymin=plow, ymax=pupp), width=1.5,
              position=position_dodge(width=2), lwd=1.3) +
  geom_point(position=position_dodge(width=2), size=1.3) + 
  facet_wrap(vars(disease), ncol=2, scales="free_y") + 
  xlab("Age (years)") + ylab("Case fatality probability")

@@

gg <- resall %>% 
  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
  dplyr::filter(! (disease=="Uterine cancer" & age > 90)) %>%
  mutate(p50 = 1 - exp(-`50%`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  ggplot(aes(x=age, y=p50, group=area)) + 
  geom_ribbon(aes(ymin=plow, ymax=pupp), 
              fill="deepskyblue") +
  geom_line() + 
  facet_grid(cols=vars(gender), rows=vars(disease), scales = "free_y", switch = "y") + 
  xlab("Age (years)") + 
  ylab("Annual case fatality probability") + 
  xlim(60,100) + 
  theme(strip.text.y.left = element_text(angle = 0))

@@

## UTRC doesn't converge, so we use the empirical Bayes results

resh %>%
  filter(disease=="Stomach cancer", gender=="Male", var=="cf") %>%
  ggplot(aes(x=age, y=`50%`, col=area, group=area)) + 
  geom_line()

resh %>% filter(gender=="Female", disease=="Stomach cancer", 
                var=="cf", age>40, age<60) %>% select(age,`50%`,Rhat) 
resnc <- resh %>% filter(Rhat > 2) 
resnc <- resh %>% filter(Rhat > 1.05) 
table(resnc$disease, resnc$gender)

@@

## Between area variability in point and interval ests, hier and nonhier models  
## Specific age and gender, compare between areas 
## Not used in paper

example_age <- 90
dat <- resall %>% filter(age==example_age, var=="cf", 
                         !(disease %in% c("Stomach cancer","Uterine cancer")),
                         #      disease %in% c("Colon and rectum cancer"),
                         model %in% c("Independent areas","Hierarchical"))
ggplot(dat, aes(x=`50%`, y=area, col=model)) +
  #   coord_cartesian(xlim=c(0.01, 0.05)) +
  facet_grid(rows=vars(disease), cols=vars(gender)) + 
  geom_point(position=ggstance::position_dodgev(height=0.4)) +
  geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
                 position=ggstance::position_dodgev(height=0.4)) +
  xlab(sprintf("Case fatality for age %s", example_age) )

## Between area variability in age curves, point ests.  CARC.  Shrinkage for oldest
## Use this in paper

pdf("../write/hier_crc.pdf", width=6, height=4)

resall %>% 
  filter(var=="cf", 
         !disease %in% c("Stomach cancer","Uterine cancer"),
         disease %in% c(
           #   "Dementia"
           "Colon and rectum cancer"
           #       "Lung cancer"
         ),
         `97.5%` < 10,
         area %in% cityregions, 
         model %in% c("Independent areas","Hierarchical")) %>%
  mutate(model = relevel(factor(model), "Independent areas")) %>%
  ggplot(aes(x=age, col=area, group=area)) +
  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), fill="gray80", col="gray80") +
  geom_line(aes(y=`50%`)) + 
  coord_cartesian(ylim=c(0, 1), 
                  xlim=c(70,  100)) +
  scale_x_continuous(breaks=seq(0,100,10)) + 
  ylab("Case fatality rate")  +
  facet_grid(cols=vars(model), rows=vars(gender)) + 
  xlab("Age (years)") + 
  theme_bw() + labs(col="")

@@

## point est curves compared between areas 
rescf <- resg %>% filter(gender=="Male", var=="cf")
rescf <- resg %>% filter(gender=="Female", var=="cf")
ggplot(rescf, aes(x=age, col=area)) +
  geom_line(aes(y=`50%`)) + 
  coord_cartesian(ylim=c(0, 1), 
                  xlim=c(50,  100)) +
  scale_x_continuous(breaks=seq(0,100,10)) + 
  ylab("Case fatality (median, 95% CrI)")  +
  facet_wrap(~disease, ncol=2) + 
  xlab("Age (years)")

@@

## Put this in the paper 
## Hazard ratio for men compared to women for case fatality 
## This is good evidence that the general pattern is common among areas 
## TODO Confirm this or otherwise by LOO CV

pdf("../../write/gender_hr.pdf")
ggplot(reshis %>% filter(age > 60), 
       aes(x=age, y=est, group=area, col=area)) + 
  geom_ribbon(aes(ymin=lower, ymax=upper), fill="gray", col=NA, alpha=0.2) + 
  geom_hline(yintercept = 1) +
  geom_line(lwd=0.8) +
  geom_line(data=resgis %>% filter(age>60), col="black", lty=2) +
  facet_wrap(~disease) +
  labs(col="Area") +
  coord_cartesian(ylim=c(0,3)) + 
  ylab("Case fatality rate ratio (men/women)") + xlab("Age (years)")
dev.off()

@@

## Lookup table for all results

100 * 9 * 18 * 2 * 3 results for each of inc and CF

TODO just save this data frame in the metahit directory. there's no standard format

TODO append the results from optimisation.  OK to have multiple results for one case 
or just save in different format, since one has mode

```{r}
app_tab <- resall %>%
  filter(var %in% c("cf","inc","rem")) %>%
  select(disease, var, gender, age, area, model, `25%`, `50%`, `97.5%`) %>%
  arrange(disease, var, gender, age, area, model)
app_tab %>% head
```

@@

Compare nonhierarchical with hierarchical model, for all diseases and outcomes.

Colon and rectum cancer is highlighted in the paper as a case where the hierarchical model makes a difference to the results, for ages 80+.   The hierarchical model gives better predictions here.

```{r}
looall <- loonh %>% mutate(model="Independent areas") %>%
  full_join(looh %>% mutate(model="Hierarchical")) %>%
  filter(!(model=="National")) %>%
  select(age, gender, elpd_loo, model, area, disease, var) %>%
  group_by(model, gender, disease, var)

looall %>% 
  summarise(eisum = sum(elpd_loo)) %>%
  pivot_wider(names_from=model, values_from=eisum) %>%
  mutate(elpd_diff = `Hierarchical` - `Independent areas`)

looall %>% 
  filter(age > 80) %>%
  summarise(eisum = sum(elpd_loo)) %>%
  pivot_wider(names_from=model, values_from=eisum) %>%
  mutate(elpd_diff = `Hierarchical` - `Independent areas`)%>%
  filter(disease=="Colon and rectum cancer")
```

Graph of how predictive ability depends on age and area, for selected disease and outcome.  Hierarchical model is better for older ages.  Age 50 data seem to be strong enough that shrinkage is not needed.  Note that age 50 parameters are also assumed to apply to younger ages.

```{r}
loosub <- looall %>% 
  filter(disease=="Colon and rectum cancer", gender=="Male", var=="mort") %>%
  ## sum ELPD by five year age group
  pivot_wider(names_from=model, values_from=elpd_loo) %>%
  mutate(elpd_diff = `Hierarchical` - `Independent areas`) %>%
  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
  group_by(area, agegroup, gender, var) %>%
  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))

ggplot(loosub, aes(x=agemid, y=elpd_diff, colour=area)) +
  geom_hline(yintercept=0, col="gray", lwd=2) +
  geom_line() + 
  geom_point() + 
  scale_x_continuous(limits=c(30,100), breaks = seq(0,100,10)) + 
  geom_text(data=data.frame(x=c(50,50), y=c(-Inf,Inf)), 
            aes(x=x, y=y), 
            inherit.aes=FALSE,
            label=c("Independent areas preferred","Hierarchical preferred"), 
            vjust=c(0,1), hjust=c(0,0)) +
  directlabels::geom_dl(aes(label=area), 
                        method=list(directlabels::dl.combine("first.points", "last.points")))

@@

Then compare hier with hier additive gender

## FIXME rerun hier with bugfix in saving loo output .  was triggered when we changed rundf
## to exclude (stmc and?) utrc

looall %>% filter(var=="mort", area=="Bristol", gender=="Male", disease=="Lung cancer", 
                  model=="Hierarchical", age==1)

# elpd : high is good. positive elpd diff favours additive gender model 
loohg <- looall %>% 
  filter(model %in% c("Hierarchical","Hierarchical joint gender")) %>%
  select(var, area, age, gender, disease, model, elpd_loo) %>%
  pivot_wider(names_from=model, values_from=elpd_loo, values_fn=length)

%>%
  mutate(elpd_diff = `Hierarchical joint gender` - `Hierarchical`) %>%
  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
  group_by(var, area, gender, disease, agegroup) %>%
  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))

loohg %>% 
  filter(disease %in% c("Ischemic heart disease","Lung cancer")) %>% 
  group_by(agemid, disease) %>%
  summarise(elpd = sum(elpd_diff)) %>%
ggplot(aes(x=agemid, y=elpd, col=disease, group=disease)) + 
  geom_hline(yintercept=0) + geom_line() + geom_point()

# ELPD all over the place for different ages 
# positive: common effects.  negative: different effects 
# Different effects for lung cancer and dem
# not transparent though. 
# IHD vs LC is still biggest contrast 
# not sure makes sense to split by age.  model defines risk as smooth function of age

loohg %>%
  group_by(var, disease) %>%
  summarise(elpd = sum(elpd_diff)) %>%
  arrange(disease) %>%
  pivot_wider(names_from=var, values_from=elpd) %>%
  filter(!is.na(inc)) %>%
  mutate(rem=ifelse(is.na(rem), 0, rem),
         overall = inc+mort+prev+rem)

>>>>

source("paper_analyses_header.r")
ymax_mort <- enframe(c("Ischemic heart disease"=0.2,"Stroke"=0.1,"Dementia"=0.2,
                       "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.02,
                       "Colon and rectum cancer"=0.02, "Stomach cancer"=0.01, 
                       "Uterine cancer"=0.002), name="disease", value="ymax_mort")

ymax_inc <- enframe(c("Ischemic heart disease"=0.1, "Stroke"=0.1,"Dementia"=0.1,
                      "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.03,
                      "Colon and rectum cancer"=0.03, "Stomach cancer"=0.01,
                      "Uterine cancer"=0.002), name="disease", value="ymax_inc")

gbdplot <- gbdplot %>%
    left_join(ymax_mort, by="disease") %>%
    left_join(ymax_inc, by="disease") %>%
    mutate(mort_upper = pmin(mort_upper, ymax_mort),
           inc_upper = pmin(inc_upper, ymax_inc),
           disease = factor(disease, levels=disease_order))
ht <- 3
pdf("../../write/mort_data.pdf", height=ht)
gbdplot$areagender <- interaction(gbdplot$area, gbdplot$gender)
ggplot(gbdplot, aes(x=age, y=mort_prob, col=area, lty=gender, group=areagender)) + 
  facet_wrap(~disease, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual mortality probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

@@

## MCMC convergence checks.   All OK with 5000 samples and the occasional model simplification
resnc <- resnh %>% filter(Rhat > 1.05)
table(resnc$disease, resnc$area, resnc$gender)

## Compare modal approximation with posterior 
## Page of plots for a single area and gender.
## Disease variation more important than area/gender variation  
## stmc, utrc use national point estimates

resopt <- readRDS("resopt_nonhier.rds") %>% 
  filter(!disease %in% c("Stomach cancer","Uterine cancer"))
resoptnat <- readRDS("resopt_nat.rds")
resoptcf <- resopt %>%
  full_join(resoptnat) %>% 
  filter(var=="cf") %>%
  mutate(p50 = 1 - exp(-`mode`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  filter(area=="Bristol" | area=="England", gender=="Female") %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten),
           disease = factor(disease, levels = disease_order)) %>%
  filter(age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(Algorithm = "Mode approximation")

@@

resp <- resall %>% 
  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
  dplyr::filter(area=="Bristol" | (area=="England" & model=="National"), gender=="Female") %>%
  mutate(Algorithm="MCMC") %>%
  full_join(resoptcf)
resp %>%
  filter(age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(par = 1 - exp(-`50%`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  ggplot(aes(x=age, y=par, col=Algorithm, lty=Algorithm)) + 
  geom_errorbar(aes(ymin=plow, ymax=pupp), width=1.5,
              position=position_dodge(width=2), lwd=1.3) +
  geom_point(position=position_dodge(width=2), size=1.3) + 
  facet_wrap(vars(disease), ncol=2, scales="free_y") + 
  xlab("Age (years)") + ylab("Case fatality probability")

@@

gg <- resall %>% 
    dplyr::filter(model %in% c("Independent areas","National"), 
                  var=="cf",
                area %in% c(cityregions,"England")) %>%
    mutate(area = forcats::fct_relevel(area, "England", after=Inf)) %>%
    dplyr::filter(! (disease=="Uterine cancer" & age > 90)) %>%
    mutate(p50 = 1 - exp(-`50%`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  ggplot(aes(x=age, y=p50, group=area, col=area)) + 
  geom_ribbon(aes(ymin=plow, ymax=pupp), 
              fill="gray80", col="gray80") +
  geom_line() + 
  facet_grid(cols=vars(gender), rows=vars(disease), scales = "free_y", switch = "y") + 
  xlab("Age (years)") + 
  ylab("Annual case fatality probability") + 
  xlim(60,100) + 
  theme(strip.text.y.left = element_text(angle = 0),
        strip.text.y = element_text(size=6),
        axis.text = element_text(size=8),
        legend.text = element_text(size=6)) + 
    labs(col="")

@@

## UTRC doesn't converge, and STMC estimates noisy, so we use national estimate results 
resnc <- resh %>% filter(Rhat > 2) 
resnc <- resh %>% filter(Rhat > 1.1) 
resnc <- resh %>% filter(Rhat > 1.05) 
table(resnc$disease, resnc$gender)

@@

## Between area variability in age curves, point ests.  All diseases

pdf("../../write/hier.pdf", width=6, height=4)

resall %>% 
    filter(var=="cf", !(disease %in% c("Uterine cancer" , "Stomach cancer")),
           area %in% cityregions, 
           gender=="Female",
           model %in% c("Independent areas","Hierarchical")) %>%
    mutate(disease = fct_recode(disease, 
                                "Colorectal cancer"="Colon and rectum cancer",
                                "IHD" = "Ischemic heart disease")) %>%
    mutate(p50 = 1 - exp(-`50%`),
           plow = 1 - exp(-`2.5%`),
           pupp = 1 - exp(-`97.5%`)) %>%
    mutate(model = relevel(factor(model), "Independent areas")) %>%
    ggplot(aes(x=age, col=area, group=area)) +
    geom_ribbon(aes(ymin=plow, ymax=pupp), fill="gray80", col="gray80") +
    geom_line(aes(y=p50)) + 
    coord_cartesian(xlim=c(70,  100), ylim=c(0,NA)) +
    scale_x_continuous(breaks=seq(0,100,10)) + 
    ylab("Annual case fatality probability")  +
    facet_grid(cols=vars(model), rows=vars(disease),
               scales="free_y", switch = "y") + 
    xlab("Age (years)") + 
    labs(col="") +
    theme(strip.text.y.left = element_text(angle = 0),
          strip.text.y = element_text(size = 6.5),
          legend.text = element_text(size = 6.5))

@@

## Point est curves compared between areas 
## Not shown in paper
rescf <- resg %>% filter(gender=="Male", var=="cf")
rescf <- resg %>% filter(gender=="Female", var=="cf")
ggplot(rescf, aes(x=age, col=area)) +
  geom_line(aes(y=`50%`)) + 
  coord_cartesian(ylim=c(0, 1), 
                  xlim=c(50,  100)) +
  scale_x_continuous(breaks=seq(0,100,10)) + 
  ylab("Case fatality (median, 95% CrI)")  +
  facet_wrap(~disease, ncol=2) + 
  xlab("Age (years)")

@@

## Put this in the paper 
## Hazard ratio for men compared to women for case fatality 
## This is good evidence that the general pattern is common among areas

pdf("../../write/gender_hr.pdf", width=6, height=4)

ggplot(reshis %>% filter(age > 60), 
       aes(x=age, y=est, group=area, col=area)) + 
  geom_ribbon(aes(ymin=lower, ymax=upper), fill="gray", col=NA, alpha=0.2) + 
  geom_hline(yintercept = 1) +
  geom_line(lwd=0.8) +
  geom_line(data=resgis %>% filter(age>60), col="black", lty=2) +
  facet_wrap(~disease) +
  labs(col="Area") +
  coord_cartesian(ylim=c(0, 3)) + 
  ylab("Case fatality rate ratio (men/women)") +
  xlab("Age (years)") + 
    theme(legend.text = element_text(size = 6.5),
          strip.text = element_text(size = 6.5))

@@

@@

Full table of model comparison statistics for all 4 relevant models for the paper

No-incidence model is best for all diseases

```{r}
looall2 <- looall %>% 
  filter(model %in% c("Independent areas",
                      "Independent areas, no incidence data",
                      "Hierarchical",
                      "Hierarchical joint gender"),
         !(disease %in% c("Uterine cancer","Stomach cancer")),
         outcome %in% c("inc","prev","mort"))

looallsum2 <- looall2 %>%
  filter(outcome %in% c("prev","mort")) %>% 
  group_by(age, area, gender, disease, model) %>%
  summarise(looic = sum(looic), .groups="drop") %>%
  mutate(outcome = "prevmort")
looallsum3 <- looall2 %>%
  group_by(age, area, gender, disease, model) %>%
  summarise(looic = sum(looic), .groups="drop") %>%
  mutate(outcome = "total")

lootab <- looall2 %>%
  full_join(looallsum2) %>%
  full_join(looallsum3) %>%
  select(outcome, area, age, gender, disease, model, looic) %>%
  pivot_wider(names_from=model, values_from=looic) %>%
  group_by(disease, outcome) %>%
  summarise(enh = sum(`Independent areas`),
            enhi = sum(`Independent areas, no incidence data`),
            eh = sum(`Hierarchical`),
            ehg = sum(`Hierarchical joint gender`), .groups=c("drop")) %>%
  mutate(enhi = ifelse(outcome %in% c("inc","total"), NA, enhi),
         disease = ifelse(outcome=="total", "", as.character(disease)),
         outcome = fct_recode(outcome, "Incidence"="inc", "Prevalence"="prev", 
                              "Mortality"="mort", "Prevalence and mortality" = "prevmort",
                              "Incidence, prevalence and mortality" ="total")) %>% 
  filter(outcome %in% c("Prevalence and mortality","Incidence, prevalence and mortality")) %>%
  mutate(across(3:6, ~ replace_na(as.character(round(.x)),"")))

lootab
write.table(lootab, row.names = FALSE, sep="  &  ", eol="\\\\\n", quote=FALSE)

@@
